Last glacial cycle glacier erosion potential in the Alps

### 2021-06-01 articles/erosion/response-referee1.md

<!-- Copyright (c) 2021, Julien Seguinot (juseg.github.io), Ian Delaney
Creative Commons Attribution-ShareAlike 4.0 International License
(CC BY-SA 4.0, http://creativecommons.org/licenses/by-sa/4.0/) -->

# Authors' response to Anonymous Referee #1

Dear Anonymous Referee #1,

Thank you very much for your positive comments and for your time spent
reviewing our work. We try to address your comments one by one below, and
highlight relevant changes made to the manuscript.

@@

> Authors decided to keep the paper short, referring most information on
> glacial model set up and resulting ice cover to the Seguinot et 2018 paper.
> This makes it no always very easy to read and needs checking in the original
> manuscript (for example, ice cover deviations model – field bases LGM
> extension).

Indeed, it is our deliberate choice to avoid paraphrasing or duplicating
content from the 2018 paper, where the preparation of the text and figures also
involved a different author team. We believe this is justified given that both
papers will be published open-access. Nonetheless, small additions were made.
Please see our responses to your technical comments below.

> The introduction gives a nice, relatively long (in relation to other
> chapters) overview on different aspects of glacial erosion and resulting
> features, and guides through problems to investigate them. Here, I don’t see
> very clear how different aspects or parts in the introduction are
> specifically addressed in this study, i.e. which problems are exactly aimed
> to be solved of this large portfolio of mentioned gaps in the understanding
> of glacial processes. I think this could be better formulated and balanced.

Our introduction is meant to revolve around two discussion points: the variable
imprint of glaciers on mountain topography, and the link between climate and
glacier erosion (somewhat addressed in the conclusions). This is followed by a
short methodological bibliography of process-based and empirical glacier
erosion laws, before we announce the scope of our work. A few key statements
were added to try and better tie these different sections.

@@

> I think the study could also well contribute to the discussion on the
> elevation distribution of the cumulative glacial erosion over several cycles
> in the Alps (high elevations vs low elevation, e.g. Valla et al., 2011). A
> plot showing the integrated glacial erosion potential over (e.g. 100 m)
> elevation bins would easy to do and eventually an enlightening supplement
> summing up Fig. 5a.

Thank you for the suggestion. Figure 5 was amended with the
distribution of cumulative erosion potential over 100 m elevation bins.
The resulting plot shows that most modelled erosion occurs below 2000 m,
with the potential erosion volumes between 1000 and 2000 m elevation.

> I would also recommend to make some regional statements on the distribution
> of the glacial erosion potential (Fig. 2). It is quite obvious that some
> tectonic massifs can readily be discriminated, e.g. the Tauern Window, Oetzal
> Crystallin complex, the Aare Massif, Mt. Rosa - Gran Paradiso and Pelvoux
> Massif. Is it the steepness of these low erodible crystallin massifs making
> the erosional potential appearing strikingly high? I think this should/could
> be discussed..

Our calculations indeed produce locally high cumulative erosion potential in
the aforementioned crystalline massifs. More accurate modelling of small-scale
glaciers would be needed to provide an accurate answer to your question. Steep
topography, though, is
likely to be part of the explanation. A more substantial discussion of the high
erosion potential values in these areas and the model limitations that hinder
interpretation was added in the "age of the glacial landscape" subsection.
This new passage reads as follow:

    The validity of the model results at high elevation is discussable.
    Crystalline massifs such as the Ecrins, Gran Paradiso, Monte Rosa, Aare,
    Ötztal and Tauern Massifs locally exhibit a strikingly high erosion
    potential. However, the computation of glacier flow velocities on such
    steep surfaces is strongly limited by the model horizontal resolution of
    1\,km, the shallow-ice glacier flow physics (Imhof et al., 2019), and
    PISM's current mass-conservation heuristics (Imhof, 2021). Besides,
    bergschrund (rimaye) processes likely to dominate interglacial cirque
    erosion at such altitudes (Sanders et al., 2012) are not captured by the
    velocity-based glacier erosion power-laws.

@@

Indeed. The increase in ice discharge was made explicit.

> Line 140 Can you be more precise what you mean by realistic? You mean because
> of the localized pattern? Maybe I am wrong but shouldn’t the erosion
> potential at least at the lake Constance overdeepening (Fig. 7) not in the
> order of hundreds of meters (cf. Preusser et al., 2010) and the best fit
> rather (b) or (c) – at least from what I read from the transect (e-h)?
> Visually (a)-(d) seem not to correspond to (e)-(h) if (a)-(d) is also
> presented in meters (annotation at the bar is missing). For example, in (b)
> the maximum erosion potential is like >>1000 m (if in meters) while in
> corresponding (f) it seems clearly lower than 1000 m.

This sentence was about the amplitude of the cumulative erosion potential, but
we were probably a bit too fast deeming one erosion law "more realistic" than
the others. We are not sure what the cumulative erosion potential of the last
glacial cycle should be. Based on the estimates of total Pleistocene glacial
relief from the suggested references, we now argue that no tested erosion law
give results in the expectable range of last glacial cycle erosion.

    With a total Pleistocene glacial relief on the order of a kilometre
    (Preusser et al., 2011; Valla et al., 2011), a cumulative glacial erosion
    for the last glacial cycle in the order of 10 to 100 m can be expected.
    However, none of the tested erosion power-laws fall within this range.
    Instead, the erosion law calibrated on tidewater glaciers (Koppes et al.,
    2015) yields cumulative erosion in the Rhine Valley in the orders of
    metres, while the three erosion laws based on terrestrial glaciers
    (Humphrey and Raymond, 1994; Herman et al., 2015; Cook et al., 2020},
    result in kilometre-scale integrated erosion potential. During the Last
    Glacial Maximum and much of the last glacial cycle, Alpine paleoglaciers
    were closer in size, slope (an important parameter as we argue in the next
    section), and climatic context to the present-day glaciers of Patagonia and
    the Antarctic Peninsula (Koppes et al., 2015) than to Franz-Joseph Glacier
    (Herman et al., 2015) and many of the glaciers included in the global
    compilation by Cook et al., (2020). This may help to explain why the
    reality appears to fall in-between the tested erosion laws.

@@

The field-based LGM outline was added on Fig. 2. Computing a percent overlap is
not straightforward. As for Fig. 4 in the 2018 paper, we use a manually
modified version of the Ehlers et al. (2011) data. The original data presents
numerous, expansive nunataks, many of which are found in accumulation area and
thus directly incompatible with the mass-balance scheme used in the 2018 paper.
However, the general eastward bias was discussed in two new sentences:

    It should be noted, however, that all runs presented here show a systematic
    bias with excessive glacier cover in the Eastern Alps and a diminished
    glacier extent in the Western Alps (Fig. 2a; further discussed in Seguinot
    et al., 2018). Thus the modelled patterns of erosion potential certainly
    includes a similar bias.

> Fig. 5 I have to confess Fig. 5a surprises me, there are really glaciers as
> low as 500 m.a.s.l (and even below) from 110 – 40 ka in the Seguinot et al.,
> 2018 model?

This is correct. It concerns only a small number of grid cells, primarily
located on the southern slope of the Alps where topography drops abruptly from
the highest peaks to deep valleys. This information was added to the plot by
hatching the regions with fewer than a hundred ice-covered grid cells per
elevation band. This
highlights that low-elevation values are not always representative, as can also
be confirmed in the new subplot showing the distribution of cumulative erosion
potential over 100 m elevation bins. The distribution of glacier cover per
elevation band over time (from which the hundred-cell contour was extracted)
will be included in a new version of the companion dataset.

@@

## References not in the manuscript:

In addition to the aforementioned changes, the Sternai et al. (2013) reference
was added in the discussion of the vertical distribution of erosion with this
new sentence:

    Over longer timescales, though, the vertical distribution of erosion rates
    also depends on the erosional modification of topography (Sternai et al.,
    2013).

Thank you again for taking the time to read our work and to give constructive
feedback in these troubled times. We apologize for the delay and will soon be
submitting a revised and, we believe, improved manuscript including the
aforementioned suggested changes.

### 2021-06-01 articles/erosion/response-referee2.md

<!-- Copyright (c) 2021, Julien Seguinot (juseg.github.io), Ian Delaney
Creative Commons Attribution-ShareAlike 4.0 International License
(CC BY-SA 4.0, http://creativecommons.org/licenses/by-sa/4.0/) -->

# Authors' response to Ian Evans

Dear Ian Evans,

Thank you very much for your and your colleague's time in preparing this
review of our manuscript. We try to address your comments one by one below, and
highlight relevant changes made to the manuscript.

@@

> The authors provide large amounts of very useful results from simulations,
> concerning the effects of varying glacier geometry on erosion potential. It
> was quite an achievement to compute this at fairly high temporal and spatial
> resolution (1 km and 50 a throughout the Alps). The conclusions are
> reasonable, and I value the results. They provide support for the
> traditional view, that high-altitude features such as cirques are eroded by
> steep, local glaciers and not under extensive icefields well above ELA. Most
> erosion is expected along major valleys (glacial troughs): the potential
> erosion is greatest low down the glaciers, with a lesser maximum at high
> altitude and a minimum between. The paper is well written and the Figures
> are all relevant and interesting. The qualitative conclusions are well
> justified.

> An interesting result, not commented on, is that from Figs. 3 and 7 i & j,
> potential erosion volume in the Alps peaks when ice cover amounts to 1.6 cm
> sea level equivalent. I would love to see a map of the glaciers at that
> stage\*, which is well below maximum but considerably more than recent ice
> cover (0.3 ± 0.1 mm s.l.e., Farinotti et al. 2019). This peak applies to the
> ‘Koppes’ and ‘Herman’ erosion laws, but not to ‘Humphrey’ and ‘Cook’ laws
> which give very broad maxima for s.l.e. 10-30 cm (Fig. 7 k & l).
>
>   [\*That stage can be approximated by pausing the first erosion
>    rates video around 99 ka and 13 ka. The three videos are strongly
>    recommended!]

Thank you for pointing this out. However, we are not confident that this result
is significant. The situation you describe occurs, for instance, during much of
MIS~5 (Fig. 2) and can be visualized in the animations (as you noted). During
such periods, glaciers are steep and only a few grid cells in width.

Increasing ice volume yields the build-up of less steep and thus slower-flowing
valley glaciers. Decreasing ice volume leads to (roughly) equally steep but
smaller glaciers. In the case of non-linear erosion laws ('Koppes' and 'Herman'
laws), both yield a decrease of erosion volume, hence the local peak in erosion
volume. However, we would rather not highlight this result. The modelled
glacier velocities for such situations are unreliable due to the limiting
horizontal resolution and the shallow-ice physics. This latter fact was
highlighted in a new passage in the section on the "age of the glacial
landscape". Also including suggestions from reviewer #1, the new passage
reads as follow:

@@

    The validity of the model results at high elevation is discussable.
    Cristalline massifs such as the Ecrins, Gran Paradiso, Monte Rosa, Aare,
    Ötztal and Tauern Massifs locally exhibit a strikingly high erosion
    potential. However, the computation of glacier flow velocities on such
    steep surfaces is strongly limited by the model horizontal resolution of
    1\,km, the shallow-ice glacier flow physics (Imhof et al., 2019), and
    PISM's current mass-conservation heuristics (Imhof, 2021). Besides,
    bergschrund (rimaye) processes likely to dominate interglacial cirque
    erosion at such altitudes (Sanders et al., 2012) are not captured by the
    velocity-based glacier erosion power-laws.

To avoid over-interpretation from future readers, this uncertainty was also
highlighted in several of our plots (including Figs. 3 and 7) using hatches
over periods of limited glacier volume, and documented in the figure captions.

@@

Thank you. Figure 4 was reworked to include a map of Rhine glacier modelled
erosion rates at 36 ka, corresponding to the last major advance phase before
the LGM and with a similar extent to the deglacial stage of 16 ka. This also
corresponds to the topographic profiles shown on Fig. 8. Thus it seems more
appropriate in hindsight.

> I find no mention of lithology, rock type, geology, resistance or erodibility
> (excluding that citing Herman on line 56). This omission of half the erosion
> equation should be mentioned; it implies that the ‘potential’ nature of the
> erosion should be emphasised repeatedly. This adds to the admitted
> limitations of lack of feedback to changing topography, absence of hydrology
> and absence of subglacial sediment accumulation or movement. Using constant
> topography might be OK for the last glaciation, as here, but would become
> increasingly unrealistic for several glaciations. Destinations of eroded
> material are not considered: deposition in temporary storage may inhibit
> erosion in parts. The omission of hydrological processes must be a major
> drawback, as Herman et al. (2011), Egholm et al. (2012a, b) and Pedersen et
> al. (2014) showed that its inclusion of subglacial meltwater, plus horizontal
> stress gradients, gave more realistic simulations of erosion. For erosion
> rate as well as for cumulative erosion, it is potential erosion that is
> calculated, rather than predicted erosion.

The above limitations were made explicit with the following added sentences in
the methods subsection on the "erosion law":

    Instead, we assume that eroded material is instantly transported out of
    the system, thus neglecting its role in shielding the bedrock from glacier
    erosion in zones of temporary storage (Preusser et al., 2010). Neither do
    we account for differences in erosion effectiveness on different
    lithologies or erosion from subglacial and interglacial hydrologic
    processes. For the aforementioned reasons, we refer to the above computed
    rates, ė, as “potential erosion rates”.

@@

> The 1 km resolution of the modelling is clearly important, so it is strange
> that the first mention is on p.11. Although it is very demanding of computer
> time, it is still (as admitted) not adequate to represent cirques (averaging
> of the order of 700 m across). Even stranger, there is no mention of the DEM
> used (it was SRTM in the 2018 paper). This information should be provided in
> section 2, p. 3-4.

The model horizontal resolutions of 1 and (for climate sensitivity runs) 2 km
is now mentioned in the methods. The input SRTM topography is also referenced
in both the methods and the caption of Fig.~2 where it is displayed in the
background. Following comments from reviewer #1, the discussion of model
limitations regarding the representation of cirque glaciers has also been
extended.

> The lateral constraints on valley glaciers are usually covered by a ‘form
> factor’: how are they handled by this version of the PISM ice sheet model –
> how has lateral drag been included?

PISM uses a combination of shallow-ice and shallow-shelf stress balances,
neither of which includes lateral stresses. Glacier sliding velocities in
troughs are governed by (the sliding law and) longitudinal stresses, which is
a limitation with respect to higher-order models. Following another comment
from reviewer #1 an additional paragraph on basal sliding uncertainties was
added, including the following sentence:

@@

> To aid interpretation of Figs 4 a-c and 7 a-d, a corresponding topographic
> map or hillshade of the Rhine area on the same scale should be provided.

In addition to the aforementioned changes to Fig. 4, a new panel was added for
the model final state at 0 ka, revealing the bedrock topography. So Fig. 4 now
has one panel during advance (36 ka), one during a maximum stage (24 ka), one
during retreat (16 ka), and one nearly ice-free (0 ka).

> Fig. 1c illustrates a very unusual cirque, with a huge lake replacing a very
> unhealthy convex glacier. Fascinating, but surely a more representative
> cirque should be used. Actually these three photos are not well integrated
> with the text (pace lines 191 & 195-198). Lauterbrunnen is an extreme
> (vertical-sided) trough example, over-used by textbooks.

The figure was moved into the discussion of the "age of the glacial landscape"
to become Fig. 8, and is not longer referred to in the intro. The photo of
Lauterbrunnental was replaced by one of Bout du Monde in the Giffre Massif, and
the caption for the cirque photo now reads: "the unusually deep mountain cirque
revealed by the current demise of Chüebodengletscher" (we don't have a much
better photo for a high-elevation cirque, and glacial lakes from retreating
cirque glaciers are a common occurrence in the Alps these days).

@@

> Fig. 5: The elevation histogram in Fig. 5a could be misleading, as it seems
> to cover the whole study area, including parts never glacier-covered. It
> should be limited to areas glacier-covered at the maximum, and preferably
> supplemented by a histogram for those covered now or at the minimum. The
> accompanying main graph (5a) relates to areas glacier-covered at each time
> slice. Re another reviewer’s comment, yes LGM glaciers did reach below 500 m:
> at L. Garda (67 m?) and the Tagliamento Glacier lobe LGM ice reached below
> 100 m. An integrated plot of total-cycle potential erosion rate (and
> volume?) at each altitude, over glacier-covered areas, would be interesting
> (as an extra Figure): can rates at high altitude, where glacial duration is
> longer, catch up with those at low altitude?  Probably not, given the
> concentration of flow down major valleys. Striping above 3000 m, presumably
> related to low surface area, suggests a switch there to 20 m or 40 m bands
> would be advisable. Caption; ‘modelled potential erosion rates’?

The figure now includes a histogram of model domain elevations, a histogram of
glacier-covered elevations, and a plot of cumulative erosion potential volume
per elevation bands of 100 m. The latter shows that, as you suspected, high
altitude rapid erosion is largely offset by the limited number of grid cells
it concerns. Instead, the bulk of the erosion potential occurs below 2000 m.

To highlight periods when low-elevation is relevant (and address the comment
from reviewer #1), we also hatched elevation bands that contain fewer than a
hundred ice-covered grid cells for a given time. We also shifted to 100-m
elevation bands on the main panel. While some detail is lost, this avoids
much of the striping (especially on the newly added contour and hatched
pattern), and we find the new plot more readable.

@@

Discussion on the magnitude of modelled erosion rates remains somewhat
speculative indeed due to the several sources of uncertainties, but the
relevant discussion text was extended and reworked:

    With a total Pleistocene glacial relief on the order of a kilometre
    (Preusser et al., 2011; Valla et al., 2011), a cumulative glacial erosion
    for the last glacial cycle in the order of 10 to 100 m can be expected.
    However, none of the tested erosion power-laws fall within this range.
    Instead, the erosion law calibrated on tidewater glaciers (Koppes et al.,
    2015) yields cumulative erosion in the Rhine Valley in the orders of
    metres, while the three erosion laws based on terrestrial glaciers
    (Humphrey and Raymond, 1994; Herman et al., 2015; Cook et al., 2020},
    result in kilometre-scale integrated erosion potential. During the Last
    Glacial Maximum and much of the last glacial cycle, Alpine paleoglaciers
    were closer in size, slope (an important parameter as we argue in the next
    section), and climatic context to the present-day glaciers of Patagonia and
    the Antarctic Peninsula (Koppes et al., 2015) than to Franz-Joseph Glacier
    (Herman et al., 2015) and many of the glaciers included in the global
    compilation by Cook et al., (2020). This may help to explain why the
    reality appears to fall in-between the tested erosion laws.

> Fig. 8a shows a flat dome at 3000 m for about the first 20 km of the Rhine.
> As there are mountains at 3630, 3583 and 3192 m around the sources, it is
> likely that   there would be steeper ice slopes up to these peaks, giving
> rather different stresses and erosion potentials. Also are the instabilities
> in Fig. 8b, near source and terminus, edge effects or related to the 1 km
> resolution?

This is correct. The "basal drag" quantity plotted on Fig. 8b is the magnitude
of a two-component vector: the basal shear stress. The basal shear stress
itself strongly depends on surface altitude gradients, but this includes both
along-flow slopes visible on the topographic profile of Fig. 8a, and the
invisible across-flow slopes. While the upper part of the topographic profile
appears smooth, there are indeed steep (bedrock and surface) slopes on either
side of the profile in this region. Because the upper part of the valley is
narrow, the glacier centre-line fluctuating over time, and the plot
interpolated between 1-km grid points, these lateral components enter the
computation of the basal shear stress magnitude in the higher reaches.

@@

    Some of the observed basal drag fluctuations in the upper part of the
    transect result from steep slopes on either side of the narrow valley.

>   The three videos are well worth watching  -in fact they clarify some
>   queries arising from the paper. Perhaps they could be captioned in the
>   ‘Video supplement’ paragraph. Some extra information on the third (bedrock
>   altitude) video could point out that there are numerous zero values not
>   shown: the zero geometric mean rates from 2300 to 3000 m around 24 ka
>   otherwise seem to conflict with the numerous positive rates plotted. Zero
>   rates presumably show where ice is frozen to the bed: their altitudinal
>   distribution might be worth discussion in the text. Above 3000 m means are
>   positive, which seems strange.

Well, it seems that you have spend a lot of time looking at our figures and
videos so let us thank you again for that. The caption of Fig. 5 mentioned "the
geometric mean of (non-zero) modelled erosion rates". We found that the
"non-zero" has no reason to be and it was removed. Sorry about the confusion
and please read on for the long explanation.

First, there are no zero values in the animation. As in the paper's Fig. 5, the
animation shows geometric means (i.e. arithmetic means in log-domain), and
geometric means can't contain zero values or they would result in a
divide-by-zero error (or a log-of-zero error). Both animation panels have a
log-scaled x-axis. While some values fall outside of the axis frame, they are
very small but not zero.

On the other hand, both Fig. 5 and the animation depict very low values for
erosion rates (especially around 24 ka). This is due to the glacier physics
embedded in PISM. In particular, the pseudo-plastic sliding law implies that
the glacier is sliding everywhere, but that sliding is infinitely small where
the basal drag is much smaller than the yield stress, including for instance
frozen-bed areas. These tiny sliding velocities yield even tinier erosion rates
(due to the power 2.34 in the erosion law). Such values are not really relevant
and they are off the colour range on Fig. 5 and off-chart in the
animation (respectively lower than 10^-9 and 10^-10 mm a^-1).

The only regions where PISM produces truly zero sliding are the regions of zero
surface slopes or zero ice thickness, where the SSA can't be solved. In
practice, this only occurs in the latter case: zero ice thickness. This is the
kind of values we have previously filtered out. Your comment led us to re-check
our calculations. We found that there is no zero-sliding values within the
ice-covered area, and thus an ice mask is sufficient to filter out undesirable
values and compute the geometric means. Hence "non-zero" was removed from the
caption.

@@

## Acknowledgements:

> I am grateful for helpful discussions with Iestyn Barr, Jeremy Ely, Matt
> Tomkins, Pippa Whitehouse, Cristina Balaban and Matt Wiecek.

> - Ian S. Evans, Durham University, U.K.

Thank you again to all of you for taking time to read our work and offer
constructive feedback in these troubled times. We hope that you will find our
answers satisfactory, apologize for the delay and will soon be submitting a
revised and, we believe, improved manuscript including the aforementioned
suggested changes.

### 2021-06-01 articles/erosion/alps-erosion.tex
Again read through the complete responses.

    Our approach does not account for feedbacks of glacier erosion onto
    bedrock topography and, then onto, ice dynamics
    \citep[e.g.,][]{Anderson.etal.2012}. Furthermore, we do not account for the
    downstream advection of debris \citep[e.g.,][]{Anderson.Anderson.2016},
    or fluvial transport of subglacial sediment \citep{Delaney.etal.2019}.
    Instead we assume that eroded material is instantly transported out of the
    system, thus neglegting its role in shielding the bedrock from glacier
    erosion in zones of temporary storage \citep{Preusser.etal.2010}. Neither
    do we account for differences in erosion effectiveness on different
    lithologies or erosion from subglacial and interglacial hydrologic
    processes. For the aforementioned reasons, we refer to the above computed
    rates, $\dot{e}$, as ``potential erosion rates''.

@@

    With a total Pleistocene glacial relief on the order of a kilometre
    \citep{Preusser.etal.2011, Valla.etal.2011}, a cumulative glacial erosion
    for the last glacial cycle in the order of 10--100\,m can be expected.
    However, none of the tested erosion power-laws fall within this range.
    Instead, the erosion law calibrated on tidewater glaciers
    \citep{Koppes.etal.2015} yields cumulative erosion in the Rhine Valley in
    the orders of metres, while the three erosion laws based on terrestrial
    glaciers \citep{Humphrey.Raymond.1994, Herman.etal.2015, Cook.etal.2020},
    result in kilometre-scale integrated erosion potential. During the Last
    Glacial Maximum and much of the last glacial cycle, Alpine paleoglaciers
    were closer in size, slope (an important parameter as we argue in the next
    section), and climatic context to the present-day glaciers of
    Patagonia and the Antarctic Peninsula \citep{Koppes.etal.2015} than to
    Franz-Joseph Glacier \citep{Herman.etal.2015} and many of the glaciers
    included in the global compilation by \citet{Cook.etal.2020}.
    This may help to explain why the reality appears to fall in-between the
    tested erosion laws.

@@

    The modelled pattern of erosion potential depends on PISM glacier physics
    and sliding model parameters. In particular, the pseudo-plastic sliding law
    exponent \citep[$q=0.25$ in][]{Seguinot.etal.2018} controls the sharpness
    of the transition between adherent and decoupled basal conditions. Recent
    ensemble validation of Antarctic Ice Sheet glacial-cycle simulations
    against geological and present-day observations \citep{Albrecht.etal.2020,
    Albrecht.etal.2020a} support a higher value of $q=0.75$, and thus a
    sliding law closer to linear, which would perhaps result in a smoother
    distribution of sliding velocity and erosion potential. Lateral stress
    gradients missing from the shallow-shelf approximation stress balance could
    also contribute to moderate sliding velocity in narrow troughs
    \citep{Herman.etal.2011, Egholm.etal.2012, Egholm.etal.2012a,
    Pedersen.etal.2014}. \todo{check these references}.

@@

    The validity of the model results at high elevation is discussable.
    Cristalline massifs such as the Ecrins, Gran Paradiso, Monte Rosa, Aare,
    \"Otztal and Tauern Massifs locally exhibit a strikingly high erosion
    potential. However, the computation of glacier flow velocities on such
    steep surfaces is strongly limited by the model horizontal resolution of
    1\,km, the shallow-ice glacier flow physics \citep{Imhof.etal.2019}, and
    PISM's current mass-conservation heuristics \citep{Imhof.2021}. Besides,
    bergschrund (rimaye) processes likely to dominate interglacial cirque
    erosion at such altitudes \citep{Sanders.etal.2012} are not captured by the
    velocity-based glacier erosion power-laws. Nevertheless,
    the lack of rapid erosion at high elevations for much of the modelled
    glacial cycle (Fig.~\ref{fig:hypsogram}) implies that the highest mountain
    cirques (e.g. Fig.~\ref{fig:landscape}c) would have formed
    over interglacial periods, when glaciers are confined to high terrain
    \citep{Barr.etal.2017, Barr.etal.2019}. Similar processes also occur on
    Himalayan glaciers and may limit erosion for the high elevation portion of
    that mountain range during cold periods as well, failing to offset uplift
    rates \citep{Harper.Humphrey.2003}. Over longer timescales, though, the
    vertical distribution of erosion rates also depends on the erosional
    modification of topography \citep{Sternai.etal.2013}.

>>>>

    Our approach does not account for feedbacks of glacier erosion onto
    bedrock topography and, then onto, ice dynamics
    \citep[e.g.,][]{Anderson.etal.2012}. Furthermore, we do not account for the
    downstream advection of debris \citep[e.g.,][]{Anderson.Anderson.2016},
    or fluvial transport of subglacial sediment \citep{Delaney.etal.2019}.
    Instead, we assume that eroded material is instantly transported out of the
    system, thus neglecting its role in shielding the bedrock from glacier
    erosion in zones of temporary storage \citep{Preusser.etal.2010}. Neither
    do we account for differences in erosion effectiveness on different
    lithologies or erosion from subglacial and interglacial hydrologic
    processes. For the aforementioned reasons, we refer to the above computed
    rates, $\dot{e}$, as ``potential erosion rates''.

@@

    With a total Pleistocene glacial relief on the order of a kilometre
    \citep{Preusser.etal.2011, Valla.etal.2011}, a cumulative glacial erosion
    for the last glacial cycle in the order of 10 to 100\,m can be expected.
    However, none of the tested erosion power-laws fall within this range.
    Instead, the erosion law calibrated on tidewater glaciers
    \citep{Koppes.etal.2015} yields cumulative erosion in the Rhine Valley in
    the orders of metres, while the three erosion laws based on terrestrial
    glaciers \citep{Humphrey.Raymond.1994, Herman.etal.2015, Cook.etal.2020},
    result in kilometre-scale integrated erosion potential. During the Last
    Glacial Maximum and much of the last glacial cycle, Alpine paleoglaciers
    were closer in size, slope (an important parameter as we argue in the next
    section), and climatic context to the present-day glaciers of
    Patagonia and the Antarctic Peninsula \citep{Koppes.etal.2015} than to
    Franz-Joseph Glacier \citep{Herman.etal.2015} and many of the glaciers
    included in the global compilation by \citet{Cook.etal.2020}.
    This may help to explain why the reality appears to fall in-between the
    tested erosion laws.

@@

    The modelled pattern of erosion potential depends on PISM glacier physics
    and sliding model parameters. In particular, the pseudo-plastic sliding law
    exponent \citep[$q=0.25$ in][]{Seguinot.etal.2018} controls the sharpness
    of the transition between adherent and decoupled basal conditions. Recent
    ensemble validation of Antarctic Ice Sheet glacial-cycle simulations
    against geological and present-day observations \citep{Albrecht.etal.2020,
    Albrecht.etal.2020a} support a higher value of $q=0.75$, and thus a
    sliding law closer to linear, which would perhaps result in a smoother
    distribution of sliding velocity and erosion potential. Lateral stress
    gradients missing from the shallow-shelf approximation stress balance could
    also contribute to moderate sliding velocity in narrow troughs
    \citep{Herman.etal.2011, Egholm.etal.2012, Egholm.etal.2012a,
    Pedersen.etal.2014}.

@@

    The validity of the model results at high elevation is discussable.
    Crystalline massifs such as the Ecrins, Gran Paradiso, Monte Rosa, Aare,
    \"Otztal and Tauern Massifs locally exhibit a strikingly high erosion
    potential. However, the computation of glacier flow velocities on such
    steep surfaces is strongly limited by the model horizontal resolution of
    1\,km, the shallow-ice glacier flow physics \citep{Imhof.etal.2019}, and
    PISM's current mass-conservation heuristics \citep{Imhof.2021}. Besides,
    bergschrund (rimaye) processes likely to dominate interglacial cirque
    erosion at such altitudes \citep{Sanders.etal.2012} are not captured by the
    velocity-based glacier erosion power-laws. Nevertheless,
    the lack of rapid erosion at high elevations for much of the modelled
    glacial cycle (Fig.~\ref{fig:hypsogram}) implies that the highest mountain
    cirques (e.g. Fig.~\ref{fig:landscape}c) would have formed
    over interglacial periods, when glaciers are confined to high terrain
    \citep{Barr.etal.2017, Barr.etal.2019}. Similar processes also occur on
    Himalayan glaciers and may limit erosion for the high elevation portion of
    that mountain range during cold periods as well, failing to offset uplift
    rates \citep{Harper.Humphrey.2003}. Over longer timescales, though, the
    vertical distribution of erosion rates also depends on the erosional
    modification of topography \citep{Sternai.etal.2013}.
